# Supplemental Data

## Data summary
This table summarizes data from `r params$year` (1 m depth only), giving the minimum, mean (average), and maximum values for each parameter. `r if(is.weekly) paste("This includes water-year (Oct-Sept) summary statistics for Secchi and temperature, which were measured year-round as part of Level I monitoring, and May-October summary statistics for all parameters. ")`To reduce biases from missing data or changes in sampling frequency, monthly means were calculated and then averaged to give an overall mean.

```{r Data summary, message=FALSE, warning=FALSE}
if(is.weekly) {
  combo.data<-bind_rows(weekly.data,L2.data)
  
  summary.L1<-lake.data.summary(combo.data,params$lake,params$year,y,wy=T,params.list="ST")
  summary.L2<-lake.data.summary(combo.data,params$lake,params$year,y,wy=F,params.list="L2")
  
  summary.table<-bind_rows(summary.L1,summary.L2) %>%
    kable(booktabs=T,
          col.names=c("Parameter","Minimum","Mean","Maximum")
    ) %>%
    kable_styling(full_width=F) %>%
    group_rows(index=c("Water-year statistics"=nrow(summary.L1),
                       "May-October statistics"=nrow(summary.L2)
    ))
  
    write_csv(bind_rows(summary.L1,summary.L2),
              sprintf("%s/Tables/%s-%s-summary.csv",
                      params$year,params$year,params$lake))
  
  } else {
  summary.L2<-lake.data.summary(L2.data,params$lake,params$year,y,wy=F,params.list="L2")
  
  summary.table<-summary.L2 %>%
    kable(booktabs=T,
          col.names=c("Parameter","Minimum","Mean","Maximum")
    ) %>%
    kable_styling(full_width=F)
  
    write_csv(summary.L2,sprintf("%s/Tables/%s-%s-alltrends.csv",
                                 params$year,params$year,params$lake))
  }

if(params$format!="word") {
  summary.table
} else {
  kable_as_image(summary.table,filename=sprintf("%s/Tables/%s-%s-summary",params$year,params$year,params$lake),file_format="jpeg")
}
```

## `r if(is.daily) paste("Hydrology: Lake level and precipitation")`
`r if(is.daily)   paste("Lake level and precipitation were recorded year-round. Bars show total weekly  precipitation, and the line shows average weekly lake level.")`

```{r Level & precip plot, fig.height=3, fig.width=6, message=FALSE, warning=FALSE}
if(is.daily) { 
  hydro<-hydro.plot(data=daily.data,lake=params$lake,year=params$year)
  
  ggsave(hydro,
         file=sprintf("%s/Plots/%s-%s-hydro.png",params$year,params$year,params$lake),
         width=6,height=3)
  
  hydro
}
```

## Water column profile
In May and August, water was collected at the mid-lake sampling station from three depths in a water-column profile: 1 m, the middle depth of the water column, and 1 m from the lake bottom.
```{r Profile table}
profile.dates<-L2.data %>%
  filter(year(Date)==params$year,
         Lake==params$lake,
         Depth>1) %>%
  group_by(Date) %>%
  summarize(nDepths=1+length(unique(Depth)))


if(nrow(profile.dates)>=1) {
profile<-L2.data %>%
  filter(Lake==params$lake,
         Date %in% profile.dates$Date,
         Parameter %in% c("Temperature",
                          "ChlorophyllA",
                          "PheophytinA",
                          "TotalNitrogen",
                          "NH3N",
                          "NO23",
                          "TotalPhosphorus",
                          "OPO4")
  ) %>%
  left_join(L2.quals,by = c("Lake", "Date", "Depth", "Parameter")) %>%
  mutate(Value=ifelse(Parameter=="TotalNitrogen",
                      round(Value,0),
                      formatC(round(Value,1),digits=1,format="f")),
         Value=as.character(Value),
         Value=ifelse(BelowMDL,sprintf("(%s)",Value),Value),
         Value=ifelse(Value=="NA","--",Value)) %>%
  select(Lake,Date,Depth,Parameter,Value) %>%
  spread(key=Parameter,value=Value) %>%
  select(Date,Depth,Temperature,ChlorophyllA,PheophytinA,
         TotalNitrogen,NH3N,NO23,
         TotalPhosphorus,OPO4)

write_csv(profile,sprintf("%s/Tables/%s-%s-profile.csv",
                          params$year,params$year,params$lake))

profile.table<-profile %>%
  mutate(DisplayDate=sprintf("%s/%s/%s",month(Date),day(Date),year(Date)),
         DisplayDate=ifelse(Depth==1,DisplayDate,"")
  )%>%
  select(DisplayDate,everything()) %>%
  select(-Date) %>%
  kable(booktabs=T,
        col.names=c("Date","Depth","Temp","Chlor","Pheo","TN","NH~3~","NO~2/3~","TP","OPO~4~"),
        align=c("l","r","r","r","r","r","r","r","r","r")
  ) %>%
  column_spec(1,bold=T) %>% 
  add_footnote(notation="symbol",
  "Parameter abbreviations are: chlorophyll-a (Chlor), pheophytin (Pheo), total nitrogen (TN), ammonia (NH3), nitrate/nitrite (NO2/3), total phosphorus (TP), orthophosphate (OPO4). Depth is in m, temperature is in °C, and all other parameters are in µg/L. Dashes indicate parameters that were not analyzed for a given sample. Values below the method detection limit (MDL) are enclosed in parentheses and have the value of the MDL substituted."
  )

if(params$format=="HTML") {
  profile.table %>%
  kable_styling(full_width=F,bootstrap_options=c("condensed","responsive"))
} else if(params$format=="word") {
profile %>%
  mutate(DisplayDate=sprintf("%s/%s/%s",month(Date),day(Date),year(Date)),
         DisplayDate=ifelse(Depth==1,DisplayDate,"")
  )%>%
  select(DisplayDate,everything()) %>%
  select(-Date) %>%
  kable(booktabs=T,
        col.names=c("Date","Depth","Temp","Chlor","Pheo","TN","NH3","NO2/3","TP","OPO4"),
        align=c("l","r","r","r","r","r","r","r","r","r")
  ) %>%
  column_spec(1,bold=T) %>% 
  add_footnote(notation="symbol",
  "Parameter abbreviations are: chlorophyll-a (Chlor), pheophytin (Pheo), total nitrogen (TN), ammonia (NH3), nitrate/nitrite (NO2/3), total phosphorus (TP), orthophosphate (OPO4). Depth is in m, temperature is in °C, and all other parameters are in µg/L. Dashes indicate parameters that were not analyzed for a given sample. Values below the method detection limit (MDL) are enclosed in parentheses and have the value of the MDL substituted."
  ) %>%
    kable_styling(latex_options="scale_down") %>%
    kable_as_image(filename=sprintf("%s/Tables/%s-%s-profile",params$year,params$year,params$lake),file_format="jpeg")
} else {
  profile.table
}


# profile.note.text<-"Parameter abbreviations are: chlorophyll-a (Chlor), pheophytin (Pheo), total nitrogen (TN), nitrate/nitrite, ammonia (NH3), total phosphorus (TP), orthophosphate (OPO4). Dashes indicate parameters that were not analyzed for a given sample. Values below the method detection limit (MDL) are enclosed in parentheses and have the value of the MDL substituted."
# 
# profile.note<-paste(text_spec(profile.note.text,font_size=12))

}

```

```{r Set up UV and alkalinity, message=FALSE, warning=FALSE, include=FALSE}

avg.alk<-L2.data %>%
  filter(year(Date)==params$year,
         Lake==params$lake,
         Parameter=="TotalAlk") %>%
  .$Value %>%
  mean(na.rm=T) %>%
  round(digits=1)

avg.UV<-L2.data %>%
  filter(year(Date)==params$year,
         Lake==params$lake,
         Parameter=="UV254") %>%
  .$Value %>%
  mean(na.rm=T) %>%
  round(digits=2)

alk.trend<-lake.trend.seasonal(data=L2.data,
                                   lake=params$lake,
                                   year=params$year,
                                   params.list="TotalAlk",
                                   overall.only=T)

has.alk.trend<-alk.trend$p<0.05 & alk.trend$slope!=0

UV.trend<-lake.trend.seasonal(data=L2.data,
                                   lake=params$lake,
                                   year=params$year,
                                   params.list="UV254",
                                   overall.only=T)

has.UV.trend<-UV.trend$p<0.05 & UV.trend$slope!=0
```


## Total alkalinity
A lake’s ability to resist acidification, also called its buffering capacity, is measured as “total alkalinity.” Lakes with total alkalinity less than 20 mg CaCO~3~ are considered sensitive to acidification. We measured total alkalinity in May and August (on profile-sampling days). In `r params$year`, the average total alkalinity was `r avg.alk` mg CaCO~3~. `r if(!is.null(t$Alkalinity)) t$Alkalinity`

The blue points and line are annual average alkalinity values for `r LakeName`. Grey points in the background are results for all other lakes in the Lake Stewardship program. 
`r if(has.alk.trend) sprintf("The dashed black line shows the long-term trend in alkalinity, with an average change of %s mg CaCO~3~ per decade.",round(alk.trend$slope*10,2))`

```{r Alkalinity plot, fig.height=3, fig.width=6, message=FALSE, warning=FALSE}
if(!is.na(avg.alk)) {

alk.plot<-lake.plot.facet(L2.data,params$lake,params$year,y,wy=F,means=T,params.list="TotalAlk")

ggsave(alk.plot,
       file=sprintf("%s/Plots/%s-%s-alkalinity.png",params$year,params$year,params$lake),
       width=6,height=3)

alk.plot
}
```

## Water color
Water color affects a lake’s water clarity (and Secchi depth). Water color is measured by shining a specific wavelength of ultraviolet light (254 nm) through a filtered water sample and measuring the percent that was absorbed. We measured UV254 absorbance in May and August (on profile-sampling days). In `r params$year`, the average UV254 absorbance was `r avg.UV`, on a scale where 0 is no absorbance (perfectly clear) and 1 is complete absorbance (perfectly opaque). `r if(!is.null(t$UV)) t$UV`

The blue points and line are annual average UV absorbance values for `r LakeName`. Grey points in the background are results for all other lakes in the Lake Stewardship program. 
`r if(has.UV.trend) sprintf("The dashed black line shows the long-term trend in UV absorbance, with an average change of %s absorbance units per decade.",round(UV.trend$slope*10,2))`

```{r UV plot, fig.height=3, fig.width=6, message=FALSE, warning=FALSE}
if(!is.na(avg.UV)) {
  
UV.plot<-lake.plot.facet(L2.data,params$lake,params$year,y,wy=F,means=T,params.list="UV254")

ggsave(UV.plot,
       file=sprintf("%s/Plots/%s-%s-UV.png",params$year,params$year,params$lake),
       width=6,height=3)

UV.plot
}
```

## Trend analysis
Below are the complete results of the seasonal-Kendall trend analysis, for all parameters at 1 m depth. In addition to the overall results, individual months' results are also shown. This can help illustrate what months are driving an overall trend. For readability, very small p-values are noted as "<0.0001".

```{r Full trend analysis}

alltrends<-lake.trend.seasonal(data=filter(bind_rows(weekly.data,L2.data),
                                           !Parameter %in% c("Microcystin","Anatoxin")),
                            lake=params$lake,
                            year=params$year,
                            params.list="all",
                            overall.only=F)

write_csv(select(alltrends,-all),
          sprintf("%s/Tables/%s-%s-alltrends.csv",params$year,params$year,params$lake))

alltrends.table.data<-alltrends %>%
  left_join(y,by="Parameter") %>%
  mutate(sig=ifelse(p<0.05 & slope!=0,T,F),
         DecadalChange=prettyNum(round(slope*10,2),format="f",digits=2,nsmall=2,drop0trailing=F), # exactly 2 digits after decimal point
         # p=scientific(p,2), # This converts p to a character -- test for p<0.05 before
         # p=prettyNum(p,format="g",digits=2,scientific=-1),
         p=ifelse(p<0.0001,"<0.0001",signif(p,1)),
         ChangeText=sprintf("%s %s",DecadalChange,units)
  ) %>%
  mutate(tablelabel=ifelse(Month=="Overall",as.character(text),NA)) %>%
  arrange(label)
  
alltrends.table<-alltrends.table.data %>%
  select(tablelabel,Month,ChangeText,p) %>%
  kable(longtable=T,booktabs=T,
        col.names=c("Parameter","Month","Change per Decade","p-value"),
        escape=F,
        align="llrc"
  ) %>%
  row_spec(which(alltrends.table.data$sig),bold=T)

when.sampled<-bind_rows(weekly.data,L2.data) %>%
  filter(Lake==params$lake,
         Parameter %in% c("Secchi","Temperature"),
         !is.na(Value)) %>%
  mutate(Year=year(Date),
         Month=month(Date)) %>%
  group_by(Parameter,Year,Month) %>%
  summarize(n=n()) %>%
  spread(key=Month,value=n)

write_csv(when.sampled,
          sprintf("%s/Tables/%s-%s-whensampled.csv",params$year,params$year,params$lake))

if(params$format=="HTML") {
  alltrends.table %>%
  kable_styling(full_width=F,bootstrap_options=c("condensed","responsive"))
} else if(params$format=="word"){
  word.table<-alltrends.table.data %>%
    mutate(p=ifelse(sig,sprintf("**%s**",p),p)) %>%
    select(tablelabel,Month,ChangeText,p) %>%
    kable(col.names=c("**Parameter**","**Month**","**Change per Decade**","**p-value**"),
          escape=F,
          align="llrc",
          format="markdown"
    )
  word.table
} else {
  alltrends.table
}

  
```

